Action Groupoids and Fractional Types

### 2016-06-09 pibackground.tex
making progress on sec 3

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background}

The main syntactic vehicle for the technical developments in this paper is a
simple language called $\Pi$ whose only computations are isomorphisms between
finite types~\cite{James:2012:IE:2103656.2103667}. After reviewing the
motivation for this language and its relevance to HoTT, we present its syntax
and semantics.

\begin{figure*}[ht]
\[\begin{array}{cc}
\begin{array}{rrcll}
\unitepl :&  \zt \oplus \tau & \iso & \tau &: \unitipl \\
\unitepr :&  \tau \oplus \zt & \iso & \tau &: \unitipr \\
\swapp :&  \tau_1 \oplus \tau_2 & \iso & \tau_2 \oplus \tau_1 &: \swapp \\
\assoclp :&  \tau_1 \oplus (\tau_2 \oplus \tau_3) & \iso & (\tau_1 \oplus \tau_2) \oplus \tau_3 
  &: \assocrp \\
\\
\unitetl :&  \ot \otimes \tau & \iso & \tau &: \unititl \\
\unitetr :&  \tau \otimes \ot & \iso & \tau &: \unititr \\
\swapt :&  \tau_1 \otimes \tau_2 & \iso & \tau_2 \otimes \tau_1 &: \swapt \\
\assoclt :&  \tau_1 \otimes (\tau_2 \otimes \tau_3) & \iso & (\tau_1 \otimes \tau_2) \otimes \tau_3 
  &: \assocrt \\
\\
\absorbr :&~ \zt \otimes \tau & \iso & \zt &: \factorzl \\
\absorbl :&~ \tau \otimes \zt & \iso & \zt &: \factorzr \\
\dist :&~ (\tau_1 \oplus \tau_2) \otimes \tau_3 & 
  \iso & (\tau_1 \otimes \tau_3) \oplus (\tau_2 \otimes \tau_3)~ &: \factor \\
\distl :&~ \tau_1 \otimes (\tau_2 \oplus \tau_3) & 
  \iso & (\tau_1 \otimes \tau_2) \oplus (\tau_1 \otimes \tau_3)~ &: \factorl 
\end{array}
& 
\begin{minipage}{0.4\textwidth}
\[\begin{array}{c}
\Rule{}
{}
{\jdg{}{}{\idiso : \tau \iso \tau}}
{}
\\
\\
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_2 \iso \tau_3}
{\jdg{}{}{c_1 \odot c_2 : \tau_1 \iso \tau_3}}
{}
\\
\\
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \oplus c_2 : \tau_1 \oplus \tau_3 \iso \tau_2 \oplus \tau_4}}
{}
\\
\\
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \otimes c_2 : \tau_1 \otimes \tau_3 \iso \tau_2 \otimes \tau_4}}
{}
\end{array}\]
\end{minipage}
\end{array}\]
Each 1-combinator $c$ has an inverse $!~c$, e.g, $!~\unitepl=\unitipl$,
$!(c_1 \odot c_2) = ~!c_2 \odot~ !c_1$, etc.
\caption{$\Pi$-combinators~\cite{James:2012:IE:2103656.2103667}
\label{pi-combinators}}
\end{figure*}

\begin{figure*}[ht]
\[\begin{array}{c}
\Rule{}
{c_1 : \tau_1 \iso \tau_2 \quad c_2 : \tau_2 \iso \tau_3 \quad c_3 : \tau_3 \iso \tau_4}
{\jdg{}{}{\assocdl : c_1 \odot (c_2 \odot c_3) \isotwo (c_1 \odot c_2) \odot c_3 : \assocdr}}
{}
\\
\\
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idldl : \idiso \odot c \isotwo c : \idldr}}
{}
\quad
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idrdl : c \odot \idiso \isotwo c : \idrdr}}
{}
\\
\\
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\rinvdl : ~! c \odot c \isotwo \idiso : \rinvdr}}
{}
\quad
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\linvdl : c ~\odot ~! c \isotwo \idiso : \linvdr}}
{}
\\
\\
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idisotwo : c \isotwo c}}
{}
\quad
\Rule{}
{c_1,c_2,c_3 : \tau_1 \iso \tau_2 \quad \alpha_1 : c_1 \isotwo c_2 \quad \alpha_2 : c_2 \isotwo c_3}
{\jdg{}{}{\transtwo ~\alpha_1 ~\alpha_2 : c_1 \isotwo c_3}}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\alpha_1 \respstwo \alpha_2 : c_1 \odot c_2 \isotwo c_3 \odot c_4}}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\respptwo ~\alpha_1 ~\alpha_2 : c_1 \oplus c_2 \isotwo c_3 \oplus c_4}}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\respttwo ~\alpha_1 ~\alpha_2 : c_1 \otimes c_2 \isotwo c_3 \otimes c_4}}
{}
\end{array}\]

@@

\caption{$\Pi$-combinators~\cite{James:2012:IE:2103656.2103667}
\label{pi-combinators2}}
\end{figure*}

\begin{figure*}[ht]
\begin{multicols}{2}
\[\begin{array}{r@{\!}lcl}
%% \evalone{\unitepl}{&(\inl{()})} \\
\evalone{\unitepl}{&(\inr{v})} &=& v \\
\evalone{\unitipl}{&v} &=& \inr{v} \\
\evalone{\unitepr}{&(\inl{v})} &=& v \\
%% \evalone{\unitepr}{&(\inr{()})} \\
\evalone{\unitipr}{&v} &=& \inl{v} \\
\evalone{\swapp}{&(\inl{v})} &=& \inr{v} \\
\evalone{\swapp}{&(\inr{v})} &=& \inl{v} \\
\evalone{\assoclp}{&(\inl{v})} &=& \inl{(\inl{v})} \\
\evalone{\assoclp}{&(\inr{(\inl{v})})} &=& \inl{(\inr{v})} \\
\evalone{\assoclp}{&(\inr{(\inr{v})})} &=& \inr{v} \\
\evalone{\assocrp}{&(\inl{(\inl{v})})} &=& \inl{v} \\
\evalone{\assocrp}{&(\inl{(\inr{v})})} &=& \inr{(\inl{v})} \\
\evalone{\assocrp}{&(\inr{v})} &=& \inr{(\inr{v})} \\
\\
\evalone{\unitetl}{&(\unitv,v)} &=& v \\
\evalone{\unititl}{&v} &=& (\unitv,v) \\
\evalone{\unitetr}{&(v,\unitv)} &=& v \\
\evalone{\unititr}{&v} &=& (v,\unitv) \\
\evalone{\swapt}{&(v_1,v_2)} &=& (v_2,v_1) \\
\evalone{\assoclt}{&(v_1,(v_2,v_3))} &=& ((v_1,v_2),v_3) \\
\evalone{\assocrt}{&((v_1,v_2),v_3)} &=& (v_1,(v_2,v_3)) \\
\end{array}\]
\[\begin{array}{r@{\!}lcl}
\evalone{\absorbr}{&(v,\_)} &=& v \\
\evalone{\absorbl}{&(\_,v)} &=& v \\
%% \evalone{\factorzl}{&()} \\
%% \evalone{\factorzr}{&()} \\
\evalone{\dist}{&(\inl{v_1},v_3)} &=& \inl{(v_1,v_3)} \\
\evalone{\dist}{&(\inr{v_2},v_3)} &=& \inr{(v_2,v_3)} \\
\evalone{\factor}{&\inl{(v_1,v_3)}} &=& (\inl{v_1},v_3) \\
\evalone{\factor}{&\inr{(v_2,v_3)}} &=& (\inr{v_2},v_3) \\
\evalone{\distl}{&(v_1,\inl{v_3})} &=& \inl{(v_1,v_3)} \\
\evalone{\distl}{&(v_2,\inr{v_3})} &=& \inr{(v_2,v_3)} \\
\evalone{\factorl}{&\inl{(v_1,v_3)}} &=& (v_1,\inl{v_3}) \\
\evalone{\factorl}{&\inr{(v_2,v_3)}} &=& (v_2,\inr{v_3}) \\
\\
\evalone{\idiso}{&v} &=& v \\
\evalone{(c_1 \odot c_2)}{&v} &=& 
  \evalone{c_2}{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inl{v})} &=& 
  \inl{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inr{v})} &=& 
  \inr{(\evalone{c_2}{v})} \\
\evalone{(c_1 \otimes c_2)}{&(v_1,v_2)} &=& 
  (\evalone{c_1}v_1, \evalone{c_2}v_2) 
\end{array}\]
\end{multicols}
\caption{\label{opsem}Operational Semantics}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%
\subsection{Reversibility}

The relevance of reversibility to HoTT is based on the following
analysis. The conventional HoTT approach starts with two, a priori, different
notions: functions and paths, and then postulates an equivalence between a
particular class of functions and paths. As illustrated above, some functions
like \AgdaFunction{not} correspond to paths. Most functions, however, are
evidently unrelated to paths. In particular, any function of type
\AgdaBound{A}~\AgdaSymbol{→}~\AgdaBound{B} that does not have an inverse of
type \AgdaBound{B}~\AgdaSymbol{→}~\AgdaBound{A} cannot have any direct
correspondence to paths as all paths have inverses. An interesting question
then poses itself: since reversible computational models --- in which all
functions have inverses --- are known to be universal computational models,
what would happen if we considered a variant of HoTT based exclusively on
reversible functions?  Presumably in such a variant, all functions --- being
reversible --- would potentially correspond to paths and the distinction
between the two notions would vanish making the univalence postulate
unnecessary. This is the precise technical idea we investigate in detail in
the remainder of the paper.

@@

The $\Pi$ family of languages is based on type isomorphisms. In the variant
we consider, the set of types $\tau$ includes the empty type $\zt$, the unit type
$\ot$, and conventional sum and product types. The values classified by these
types are the conventional ones: $\unitv$ of type $\ot$, $\inl{v}$ and $\inr{v}$ for
injections into sum types, and $(v_1,v_2)$ for product types:
\[\begin{array}{lrcl}
(\textrm{Types}) & 
  \tau &::=& \zt \alt \ot \alt \tau_1 \oplus \tau_2 \alt \tau_1 \otimes \tau_2 \\
(\textrm{Values}) & 
  v &::=& \unitv \alt \inl{v} \alt \inr{v} \alt (v_1,v_2) \\
(\textrm{1-combinators}) & 
  c &:& \tau_1 \iso \tau_2 ~ [\textit{see Fig.~\ref{pi-combinators}}] \\
(\textrm{2-combinators}) &
  \alpha &:& c_1 \isotwo c_2 \mbox{~where~} c_1, c_2 : \tau_1 \iso \tau_2 \\
  & && [\textit{see Fig.~\ref{pi-combinators2}}]
\end{array}\]
The interesting syntactic category of $\Pi$ is that of \emph{combinators}
which are witnesses for type isomorphisms $\tau_1 \iso \tau_2$. They consist
of base combinators (on the left side of Fig.~\ref{pi-combinators}) and
compositions (on the right side of the same figure). Each line of the figure
on the left introduces a pair of dual constants\footnote{where $\swapp$ and
$\swapt$ are self-dual.} that witness the type isomorphism in the
middle. This set of isomorphisms is known to be
complete~\cite{Fiore:2004,fiore-remarks} and the language is universal for
hardware combinational
circuits~\cite{James:2012:IE:2103656.2103667}.\footnote{If recursive types
and a trace operator are added, the language becomes Turing
complete~\cite{James:2012:IE:2103656.2103667,rc2011}. We will not be
concerned with this extension in the main body of this paper but it will be
briefly discussed in the conclusion.}

@@

We give an operational semantics for $\Pi$.  Operationally, the
semantics consists of a pair of mutually recursive evaluators that
take a combinator and a value and propagate the value in the
``forward'' $\triangleright$ direction or in the
``backwards''~$\triangleleft$ direction. We show the complete forward
evaluator in Fig.~\ref{opsem}; the backwards evaluator differs in
trivial ways.

Every 1-combinator has an \emph{order} $n$: applying this combinator $n$ times is equivalent to the identity

A couple of examples: compute order

The type $(\ot \oplus \ot) \oplus \ot$ has cardinality 3 and hence has
6 automorphisms on it:

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Programming with Equivalences}

The main syntactic vehicle for the technical developments in this
paper is the language $\Pi$ whose only computations are isomorphisms
between finite types and equivalences between these
isomorphisms~\cite{James:2012:IE:2103656.2103667,Carette2016}. We
present the syntax and operational and denotational semantics of
the parts of the language relevant to our development.

\begin{figure*}[ht]
\[\begin{array}{cc}
\begin{array}{rrcll}
\unitepl :&  \zt \oplus \tau & \iso & \tau &: \unitipl \\
\unitepr :&  \tau \oplus \zt & \iso & \tau &: \unitipr \\
\swapp :&  \tau_1 \oplus \tau_2 & \iso & \tau_2 \oplus \tau_1 &: \swapp \\
\assoclp :&  \tau_1 \oplus (\tau_2 \oplus \tau_3) & \iso & (\tau_1 \oplus \tau_2) \oplus \tau_3 
  &: \assocrp \\
\\
\unitetl :&  \ot \otimes \tau & \iso & \tau &: \unititl \\
\unitetr :&  \tau \otimes \ot & \iso & \tau &: \unititr \\
\swapt :&  \tau_1 \otimes \tau_2 & \iso & \tau_2 \otimes \tau_1 &: \swapt \\
\assoclt :&  \tau_1 \otimes (\tau_2 \otimes \tau_3) & \iso & (\tau_1 \otimes \tau_2) \otimes \tau_3 
  &: \assocrt \\
\\
\absorbr :&~ \zt \otimes \tau & \iso & \zt &: \factorzl \\
\absorbl :&~ \tau \otimes \zt & \iso & \zt &: \factorzr \\
\dist :&~ (\tau_1 \oplus \tau_2) \otimes \tau_3 & 
  \iso & (\tau_1 \otimes \tau_3) \oplus (\tau_2 \otimes \tau_3)~ &: \factor \\
\distl :&~ \tau_1 \otimes (\tau_2 \oplus \tau_3) & 
  \iso & (\tau_1 \otimes \tau_2) \oplus (\tau_1 \otimes \tau_3)~ &: \factorl 
\end{array}
& 
\begin{minipage}{0.4\textwidth}
\[\begin{array}{c}
\Rule{}
{}
{\jdg{}{}{\idiso : \tau \iso \tau}}
{}
\\
\\
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_2 \iso \tau_3}
{\jdg{}{}{c_1 \odot c_2 : \tau_1 \iso \tau_3}}
{}
\\
\\
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \oplus c_2 : \tau_1 \oplus \tau_3 \iso \tau_2 \oplus \tau_4}}
{}
\\
\\
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \otimes c_2 : \tau_1 \otimes \tau_3 \iso \tau_2 \otimes \tau_4}}
{}
\end{array}\]
\end{minipage}
\end{array}\]
Each 1-combinator $c$ has an inverse $!~c$, e.g, $!~\unitepl=\unitipl$,
$!(c_1 \odot c_2) = ~!c_2 \odot~ !c_1$, etc.
\caption{$\Pi$ 1-combinators~\cite{James:2012:IE:2103656.2103667}
\label{pi-combinators}}
\end{figure*}

\begin{figure*}[ht]
\[\begin{array}{c}
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idisotwo : c \isotwo c}}
{}
~
\Rule{}
{c_1,c_2,c_3 : \tau_1 \iso \tau_2 \quad \alpha_1 : c_1 \isotwo c_2 \quad \alpha_2 : c_2 \isotwo c_3}
{\jdg{}{}{\transtwo ~\alpha_1 ~\alpha_2 : c_1 \isotwo c_3}}
{}
~
\Rule{}
{c_1 : \tau_1 \iso \tau_2 \quad c_2 : \tau_2 \iso \tau_3 \quad c_3 : \tau_3 \iso \tau_4}
{\jdg{}{}{\assocdl : c_1 \odot (c_2 \odot c_3) \isotwo (c_1 \odot c_2) \odot c_3 : \assocdr}}
{}
\\
\\
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idldl : \idiso \odot c \isotwo c : \idldr}}
{}
~
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idrdl : c \odot \idiso \isotwo c : \idrdr}}
{}
~
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\rinvdl : ~! c \odot c \isotwo \idiso : \rinvdr}}
{}
~
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\linvdl : c ~\odot ~! c \isotwo \idiso : \linvdr}}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\alpha_1 ~\respstwo~ \alpha_2 : c_1 \odot c_2 \isotwo c_3 \odot c_4}}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\respptwo ~\alpha_1 ~\alpha_2 : c_1 \oplus c_2 \isotwo c_3 \oplus c_4}}
{}
~
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\respttwo ~\alpha_1 ~\alpha_2 : c_1 \otimes c_2 \isotwo c_3 \otimes c_4}}
{}
\end{array}\]

@@

\caption{$\Pi$ 2-combinators (excerpt)~\cite{Carette2016}
\label{pi-combinators2}}
\end{figure*}

\begin{figure*}[ht]
\begin{multicols}{2}
\[\begin{array}{r@{\!}lcl}
%% \evalone{\unitepl}{&(\inl{()})} \\
\evalone{\unitepl}{&(\inr{v})} &=& v \\
\evalone{\unitipl}{&v} &=& \inr{v} \\
\evalone{\unitepr}{&(\inl{v})} &=& v \\
%% \evalone{\unitepr}{&(\inr{()})} \\
\evalone{\unitipr}{&v} &=& \inl{v} \\
\evalone{\swapp}{&(\inl{v})} &=& \inr{v} \\
\evalone{\swapp}{&(\inr{v})} &=& \inl{v} \\
\evalone{\assoclp}{&(\inl{v})} &=& \inl{(\inl{v})} \\
\evalone{\assoclp}{&(\inr{(\inl{v})})} &=& \inl{(\inr{v})} \\
\evalone{\assoclp}{&(\inr{(\inr{v})})} &=& \inr{v} \\
\evalone{\assocrp}{&(\inl{(\inl{v})})} &=& \inl{v} \\
\evalone{\assocrp}{&(\inl{(\inr{v})})} &=& \inr{(\inl{v})} \\
\evalone{\assocrp}{&(\inr{v})} &=& \inr{(\inr{v})} \\
\\
\evalone{\unitetl}{&(\unitv,v)} &=& v \\
\evalone{\unititl}{&v} &=& (\unitv,v) \\
\evalone{\unitetr}{&(v,\unitv)} &=& v \\
\evalone{\unititr}{&v} &=& (v,\unitv) \\
\evalone{\swapt}{&(v_1,v_2)} &=& (v_2,v_1) \\
\evalone{\assoclt}{&(v_1,(v_2,v_3))} &=& ((v_1,v_2),v_3) \\
\evalone{\assocrt}{&((v_1,v_2),v_3)} &=& (v_1,(v_2,v_3)) \\
\end{array}\]
\[\begin{array}{r@{\!}lcl}
\evalone{\absorbr}{&(v,\_)} &=& v \\
\evalone{\absorbl}{&(\_,v)} &=& v \\
%% \evalone{\factorzl}{&()} \\
%% \evalone{\factorzr}{&()} \\
\evalone{\dist}{&(\inl{v_1},v_3)} &=& \inl{(v_1,v_3)} \\
\evalone{\dist}{&(\inr{v_2},v_3)} &=& \inr{(v_2,v_3)} \\
\evalone{\factor}{&\inl{(v_1,v_3)}} &=& (\inl{v_1},v_3) \\
\evalone{\factor}{&\inr{(v_2,v_3)}} &=& (\inr{v_2},v_3) \\
\evalone{\distl}{&(v_1,\inl{v_3})} &=& \inl{(v_1,v_3)} \\
\evalone{\distl}{&(v_2,\inr{v_3})} &=& \inr{(v_2,v_3)} \\
\evalone{\factorl}{&\inl{(v_1,v_3)}} &=& (v_1,\inl{v_3}) \\
\evalone{\factorl}{&\inr{(v_2,v_3)}} &=& (v_2,\inr{v_3}) \\
\\
\evalone{\idiso}{&v} &=& v \\
\evalone{(c_1 \odot c_2)}{&v} &=& 
  \evalone{c_2}{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inl{v})} &=& 
  \inl{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inr{v})} &=& 
  \inr{(\evalone{c_2}{v})} \\
\evalone{(c_1 \otimes c_2)}{&(v_1,v_2)} &=& 
  (\evalone{c_1}v_1, \evalone{c_2}v_2) 
\end{array}\]
\end{multicols}
\caption{\label{opsem}$\Pi$ operational semantics}
\end{figure*}

@@

The $\Pi$ family of languages is based on type isomorphisms. In the
variant we consider, the set of types $\tau$ includes the empty
type~$\zt$, the unit type $\ot$, and sum $\oplus$ and product
$\otimes$ types. The values classified by these types are the
conventional ones: $\unitv$ of type~$\ot$, $\inl{v}$ and $\inr{v}$ for
injections into sum types, and $(v_1,v_2)$ for product types. The
language has two other syntactic categories of programs to be
described in detail:
\[\begin{array}{lrcl}
(\textrm{Types}) & 
  \tau &::=& \zt \alt \ot \alt \tau_1 \oplus \tau_2 \alt \tau_1 \otimes \tau_2 \\
(\textrm{Values}) & 
  v &::=& \unitv \alt \inl{v} \alt \inr{v} \alt (v_1,v_2) \\
(\textrm{1-combinators}) & 
  c &:& \tau_1 \iso \tau_2 ~ [\textit{see Fig.~\ref{pi-combinators}}] \\
(\textrm{2-combinators}) &
  \alpha &:& c_1 \isotwo c_2 \mbox{~where~} c_1, c_2 : \tau_1 \iso \tau_2 \\
  & && [\textit{see Fig.~\ref{pi-combinators2}}]
\end{array}\]

Both classes of programs, 1-combinators $c$ and
2-combinators~$\alpha$, denote \emph{equivalences} in the HoTT
sense. The elements $c$ of 1-combinators denote type isomorphisms. The
elements $\alpha$ of 2-combinators denote equivalences between these
type isomorphisms. Using the 1-combinators, it is possible to write
any reversible boolean function and hence encode arbitrary boolean
functions by a technique that goes back to \citet{Toffoli:1980}. The
2-combinators provide a layer of programs that computes semantic
preserving transformations of 1-combinators. As a small example, let
us abbreviate $\ot \oplus \ot$ as $\mathsf{Bool}$ and examine two
possible implementations of boolean negation. The first directly uses
the primitive combinator
$\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to swap the
two values of type $\mathsf{Bool}$; the second use three consecutive
copies of $\swapp$ to achieve the same effect:
\[\begin{array}{rcl}
\mathsf{not_1} &=& \swapp \\
\mathsf{not_2} &=& (\swapp \odot \swapp) \odot \swapp 
\end{array}\]
We can write a 2-combinator whose \emph{type} is $\mathsf{not_2}
\isotwo \mathsf{not_1}$:
\[
\transtwo~(\idldl ~\respstwo~ \idisotwo)~\idldl
\]
which not only shows the equivalence of the two implementations of
negation but also shows \emph{how} to transform one to the other.

Fig.~\ref{pi-combinators} lists all the 1-combinators which consist of
base combinators (on the left) and compositions (on the right). Each
line of the base combinators introduces a pair of dual
constants\footnote{where $\swapp$ and $\swapt$ are self-dual.} that
witness the type isomorphism in the middle. This set of isomorphisms
is known to be sound and
complete~\cite{Fiore:2004,fiore-remarks}. Fig.~\ref{pi-combinators2}
lists a few of the 2-combinators that we use in this paper. Each
2-combinator relates two 1-combinators of the same type and witnesses
their equivalence. The 2-equivalences behave as expected with respect
to inverses of 1-combinators as shown below.

@@

We give an operational semantics for $\Pi$.  Operationally, the
semantics consists of a pair of mutually recursive evaluators that
take a combinator and a value and propagate the value in the forward
direction~$\triangleright$ or in the backward
direction~$\triangleleft$. We show the complete forward evaluator in
Fig.~\ref{opsem}; the backward evaluator is easy to infer.

As an example, let $\mathbb{3}$ abbreviate the type
$(\ot \oplus \ot) \oplus \ot$ and consider the following six
combinators of type $\mathbb{3} \iso \mathbb{3}$:

@@

There are three values of type $\mathbb{3}$ which are
$\inl{\inl{\unitv}}$, $\inl{\inr{\unitv}}$, and
$\inr{\unitv}$. Evaluting $a_2$ on each of these values, we get:
\[\begin{array}{rcl}
\evalone{(\swapp\oplus\idiso)}{\inl{\inl{\unitv}}} &=& \inl{\evalone{\swapp}{\inl{\unitv}}} \\
&=& \inl{\inr{\unitv}} \\
\\
\evalone{(\swapp\oplus\idiso)}{\inl{\inr{\unitv}}} &=& \inl{\evalone{\swapp}{\inr{\unitv}}} \\
&=& \inl{\inl{\unitv}} \\
\\
\evalone{(\swapp\oplus\idiso)}{\inr{\unitv}} &=& \inr{\evalone{\idiso}{\unitv}} \\
&=& \inr{\unitv}
\end{array}\]

Every 1-combinator has an \emph{order} $n$: applying this combinator
$n$ times is equivalent to the identityA couple of examples: compute order
The type $(\ot \oplus \ot) \oplus \ot$ has cardinality 3 and hence has
6 automorphisms on it:
